The promise of community driven preprints in ecology and evolution

### 2024-04-18 ms/ms.qmd
Updated MS. Close, but need more refs now

---
title: "Encouraging the use of community driven preprints in ecology and evolution"
bibliography: ../bib/refs.bib
csl: ../bib/nature.csl
format:
  docx:
    reference-doc: ../bib/template.docx
  pdf:
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])`.  *EcoEvoRxiv* accepts both preprints and postprints. Postprints are articles that are already published but are the final version of the published paper before typesetting for the research journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a preprint policy that recognised preprints - though most do). For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints in our 'demographics' analysis.

@@

After removing `r dim(dupl)[1]` duplicate titles – suggesting a few authors created multiple submissons for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55-60 preprints published each month in the last two years (@fig-summary A). Articles covered all major taxomic groups, with the most common being plants (`r sum_taxa$perc[1]`%), vertebrates (`r sum_taxa$perc[6]`%), and invertebrates (`r sum_taxa$perc[5]`%) (@fig-summary B). A total of `r length(unique(countries$country))-1` countries have used *EcoEvoRxiv* as a preprint server. Unsurprisingly, countries in North America and Europe preprint more exstensively. Countries in Africa, Central America and parts of Asia are still slow to use *EcoEvoRxiv* (@fig-summary C). Overall, 90% of the articles come from just `r nrow(most_use)-1` countries.

@@

Given that *EcoEvoRxiv* is a community driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses. Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem. As a result, articles on *EcoEvoRxiv* are more diverse in nature than other preprint servers; nonetheless, typical research articles (`r sum_types[sum_types$type == "research article",]$perc`%)are still the most common type of preprint, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) (@fig-summary).

Reports, parcticularly from government and industry research areas exist but are not yet common on *EcoEvoRxiv*. We would encourage authors to consider posting these types of articles to ensure they are openly accessible and more easily found. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) to be a place where their reports can be lodged to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these reports meet the IUCN requirements.

@@

*EcoEvoRxiv* is the only preprint server that has worked hard to break down language barriers to scientific communication; currently accepting English, Spanish, Portuguese and French manuscripts. Such initiatives are incredible important if we are to Multilingual initiatives have been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in our other accepted langauges (e.g., French and Portuguese). Part of the challenge in getting authors to submit multilingual preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries; cultural differences in the perception of preprints and a strong reliance on traditional publishing models, which typically favour publishing in English.

@@

Our data driven approach has allowed us to better understand some of the key elements of uncertainty surrounding preprints which provides us with opportunties to clarify our perspective and recommendations. First, authors need not be afraid of sharing data and code early in the publication process. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings. If authors are worried about data being used unintentially, clear information surrounding its re-use and withholding some data or meta-data needed to understand the dataset may be options. Second, authors need to be aware of the benefits of posting preprints and postprints. Posting preprints can help establish precedence of research findings, ensure that research is open access, and help authors receive feedback on their research before it is published. It is notable that discussions around a preprints findings are still rare and its clear that we as a community need to provide better platforms that document discussions around a preprints findings. Third, authors need to be aware of the benefits of using community driven preprint servers. Community driven preprint servers are capable of setting their own agenda and are driven by the needs and disires of the community they would like their findings to be shared with. Supporting these initiatives should be a priority for all researchers. Lastly, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to journals deemed suitable for the work. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recomendations when deciding to publish a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process.

Some authors have been frustrated by preprints and publications being disconnected in the digital world, with many instances where preprints and publications are not correctly recognised as the same paper that has developed over the peer review process. It was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%) and this was the major contributer to preprints and published articles not being matched. As such, we would recommend that authors make sure that, when preprints end up getting published, that they update preprints with the publication DOI when it's available. Particularly if their title has changed. This is quite easy for authors to do in *EcoEvoRxiv* and it will ensure that the preprint is linked to the published article and the citations are merged and the published paper is ultimately open access, allowing authors to save open access fees.

Despite the early successes of the new initiatives taken by *EcoEvoRxiv* it is still clear that much work needs to be done to improve our communities understanding of preprints (and postprints). We view this perspective piece as one small step in achieving that goal.

>>>>

---
title: "The promise of community driven preprints in ecology and evolution"
bibliography: ../bib/refs.bib
csl: ../bib/nature.csl
format:
  docx:
    reference-doc: ../bib/template.docx
  pdf:
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])`.  *EcoEvoRxiv* accepts both preprints and postprints. Postprints are articles that are already published but are the final version of the published paper before typesetting for the research journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a preprint policy that recognises preprints - though most do). For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints in our 'demographics' analysis.

@@

After removing `r dim(dupl)[1]` duplicate titles – suggesting a few authors created multiple submissons for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55-60 preprints published each month in the last two years ([@fig-summary]A). A total of `r length(unique(countries$country))-1` countries have used *EcoEvoRxiv* as a preprint server. Unsurprisingly, countries in North America and Europe preprint more exstensively. Countries in Africa, Central America and parts of Asia are still slow to use *EcoEvoRxiv* ([@fig-summary]B). Overall, 90% of the articles come from just `r nrow(most_use)-1` countries. Articles covered all major taxomic groups, with the most common being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%) and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

@@

Given that *EcoEvoRxiv* is a community driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses. Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem. As a result, articles on *EcoEvoRxiv* are more diverse in nature than other preprint servers; nonetheless, typical research articles (`r sum_types[sum_types$type == "research article",]$perc`%) are still the most common type of preprint, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) (@fig-summary).

Reports, parcticularly from government and industry research areas exist but are not yet common on *EcoEvoRxiv*. We would encourage authors to consider posting these types of articles to ensure they are openly accessible and more easily found. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) to be a place where their reports can be lodged to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these reports meet the IUCN requirements. Such reports often contain important data that can be leveraged for research syntheses and meta-analyses.

@@

*EcoEvoRxiv* is the only preprint server that has worked hard to break down language barriers to scientific communication; currently accepting English, Spanish, Portuguese and French manuscripts. Such initiatives are incredibly important if we are to begin filling global voids of scientitic knowledge, Multilingual initiatives have been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in our other accepted langauges (e.g., French and Portuguese). Part of the challenge in getting authors to submit multilingual preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries; cultural differences in the perception of preprints and a strong reliance on traditional publishing models, which typically favour publishing in English.

@@

Our data driven approach has allowed us to better understand some of the key elements of uncertainty surrounding preprints and identify shortcomings and areas needing greater clarity to authors. A number of key insights can be gleaned upon from an analysis of the first *EcoEvoRxiv* preprints.

- First, authors need not be afraid of sharing data and code early in the publication process. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings. If authors are worried about data being used unintentially, clear information surrounding its re-use and withholding some data or meta-data needed to understand the dataset may be options. 
- Second, authors need to be aware of the benefits of posting preprints and postprints. Posting preprints is expected to help: 1) establish precedence of research findings; 2) ensure that research is open access and 3) authors receive feedback on their research before it is published. While it is clear points 1 and 2 have already been achived discussions around a preprints findings are still rare. It is clear that we, as a community, need to provide better platforms that document discussions around a preprint findings. Such discussions are useful for the authors to help improve their work and communicate their findings more effectively (when done constructively, of course).
- Third, authors need to be aware of the benefits of using community driven preprint servers. Community driven preprint servers are capable of setting their own agenda and are driven by the needs and disires of the community they would like their findings to be shared with. Supporting these initiatives should be a priority for all researchers. 
- Lastly, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to journals deemed suitable for the work. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recomendations when deciding to publish a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process.

While most preprints seemlessly get connected with their published version and merged together often they can remain disconnected making it unclear how to associate the preprint and published version, creating confusion in the findings and leading to frustration from authors. Cross-refererencing problems do occur and are the major reason for why preprints are not always connected automatically. Thanksfully, the reasons are fairly well understood and easily rectified. They often result from a mismatch between titles or authors in preprint and published metadata. In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%) and this was the major contributer to preprints and published articles not being matched automatically in Google Scholar during our data collection. As such, we would recommend that authors make sure that, when preprints end up getting published, that they update their preprint with the publication DOI when it's available. Particularly if their title has changed. This is quite easy for authors to do in *EcoEvoRxiv* and it will ensure that the preprint is linked to the published article and the citations are merged and the published paper is ultimately open access, allowing authors to save open access fees.

Despite the early successes of the new initiatives taken by *EcoEvoRxiv* it is still clear that much work needs to be done to improve our communities understanding of preprints (and postprints). We view this perspective piece as one small step in achieving that goal. We believe that the future of preprints is bright and community driven initiatives, such as *EcoEvoRxiv*, are set to play a key role in the future of scientific publishing.
